Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-17 expected_and_actual_counts_ratio/notebooks/expected_and_actual_counts_ratio.Rmd
Modify plots describing the data to only use spike mutations and fix coloring of volcano plots

```{r read_data}
# Read data using access to the Fred Hutch server.
SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)

@@

# Identify the two clades that will be included in comparisons.
clades <- c("22A","22B")

@@

# consider entire gene column
# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
```

@@

After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visual the proportion of each mutation for each clade. The number of mutations is normalized by the expected count.

```{r stacked_nucleotide, echo=FALSE}
stacked_nt <- nt_group
stacked_nt$mutation_type <- paste(str_sub(nt_group$nt_mutation,1,1), ".", str_sub(nt_group$nt_mutation,-1,-1), sep="")

@@

In the following amino acid mutation plot, individual plots indicate the starting amino acid (i.e., clade founder) and stacked bars within each plot indicate what that amino acid mutated to.

```{r stacked_amino_acid, echo=FALSE}
stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)

stacked_aa <- aa_group
stacked_aa$clade_founder <- str_sub(aa_group$clade_founder_aa,1,1)
stacked_aa$mutant <- str_sub(aa_group$mutant_aa,1,1)

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder, ncol=7)+
  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
  scale_x_discrete(guide=guide_axis(n.dodge=5))+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))

# Save and load the variable because calculating p-values may take a long time.
#save(p_values, file=paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
load(paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
```

@@

```{r volcano_corrected, echo=FALSE}
# After dropping P = 1, plot the corrected p-values.
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
if("gene" %notin% colnames(volcano)){
  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep=";"))
}
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))

p <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()

p + theme(legend.position="none")
```

```{r volcano_corrected_legend}
p + theme(legend.position="bottom")
```

>>>>

```{r read_data}
# Read data using access to the Fred Hutch server.
SARS_mutation <- read.csv("../../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)

@@

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22B")

@@

# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
```

@@

After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visual the proportion of each mutation for each clade. The number of mutations is normalized by the expected count. To simplify the plot, only the nucleotide mutations in the spike protein are plotted. Previous studies have shown that this protein is important for SARS-CoV-2 evolution.

```{r stacked_nucleotide, echo=FALSE}
stacked_nt <- nt_group[which(nt_group$gene=="S"),]
stacked_nt$mutation_type <- paste(str_sub(stacked_nt$nt_mutation,1,1), ".", str_sub(stacked_nt$nt_mutation,-1,-1), sep="")

@@

In the following amino acid mutation plot, individual plots indicate the starting amino acid (i.e., clade founder) and stacked bars within each plot indicate what that amino acid mutated to. Again, only the amino acid mutations in the spike protein are plotted.

```{r stacked_amino_acid, echo=FALSE}
stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)
stacked_aa <- aa_group[which(aa_group$gene=="S"),]

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant_aa))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder_aa, ncol=5)+
  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
  scale_x_discrete(guide=guide_axis(n.dodge=5))+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))

# Save and load the variable because calculating p-values may take a long time.
#save(p_values, file=paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
load(paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
```

@@

```{r volcano_corrected, echo=FALSE, fig.height=10}
# After dropping P = 1, plot the corrected p-values.
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
if("gene" %notin% colnames(volcano)){
  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep="_"))
}
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))

ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()+
  theme(legend.position="bottom")
```

Plot the same volcano plot but color only the points where the mutation is in the spike protein for one or both clades.

```{r volcano_corrected_spike, echo=FALSE}
spike <- volcano$gene[grep("S", volcano$gene)]
volcano$spike <- ifelse(volcano$gene %in% spike, volcano$gene, "non-spike")
non_spike <- which(levels(as.factor(volcano$spike))=="non-spike")
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$spike)))
color[non_spike] <- "grey"

ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=spike))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()
```
